---
title: "low_tpm_gene_across_tissue"
author: "shaikh mohammad kounen"
date: "2026-07-28"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

library(readr)
library(dplyr)
library(ggplot2)
library(scales)
```


## Low tpm genes across tissue

**load data**

```{r}
cat_species <- read_csv("C:/Users/Lenovo/Documents/mered data/cat/cat.csv",
show_col_types = FALSE
                        
        )
```

- now we have to filter data from merge file into gene who have  tpm<1.
```{r}

cat_low_tpm <- cat_species %>%
  filter(TPM < 1)
```

- now we get no of total low tpm gene.
```{r}
cat("Total number of rows with TPM < 1:", nrow(cat_low_tpm))
```

## Tissue with most no of Low tpm genes

```{r}
low_tissue <- cat_low_tpm %>%
  group_by(`Anatomical entity name`) %>%
  summarise(
    Low_TPM_Genes = n_distinct(`Gene ID`),
    .groups = "drop"
  ) %>%
  arrange(desc(Low_TPM_Genes))
print(low_tissue)
```
**from above tabe we can conclude that liver has higest no low tpm genes and embyonic head have lowest**

## bar-plot on no of low tpm gene across tissue

```{r}


ggplot(low_tissue,
       aes(x = reorder(`Anatomical entity name`, Low_TPM_Genes),
           y = Low_TPM_Genes,
           fill = Low_TPM_Genes)) +

  geom_col(width = 0.8) +

  coord_flip() +

  geom_text(aes(label = Low_TPM_Genes),
           hjust = 1.2,
          color = "white",
            size = 3.8) +

  scale_fill_gradient(
    low = "#90CAF9",
    high = "#0D47A1"
  ) +

  labs(
    title = "Number of Low-Expression Genes Across Cat Tissues",
    subtitle = "Unique genes with TPM < 1 in each tissue",
    x = "Tissue",
    y = "Number of Low-Expression Genes"
  ) +

  theme_minimal(base_size = 14) +

  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    axis.title = element_text(face = "bold"),
    axis.text.y = element_text(size = 9),
    axis.text.x = element_text(size = 10),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )
```

### conclusion
**Low-expression gene analysis (TPM < 1) revealed clear differences among cat tissues. The liver exhibited the highest number of low-expression genes (23,968), whereas the embryonic head had the fewest (5,669). These results demonstrate that the abundance of lowly expressed genes varies across tissues**
